Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-31 papers/wdmerger_I/paper.tex
Flip the sign of the Poisson equation. Also, a brief summary to the gravity-hydro coupling section.

We solve the Poisson equation for self-gravity for our problem,
\begin{equation}
  \nabla^2 \Phi(\mathbf{x}) = -4\pi G\, \rho(\mathbf{x}),
\end{equation}
where $\Phi$ is the gravitational potential, $G$ is the gravitational
constant, and $\rho$ is the mass density. We note that the sign
convention used in CASTRO is opposite to that commonly seen in the
physics literature, so that $\Phi$ is positive everywhere. The
solution of this equation in CASTRO is described in \cite{castro}, and
consists of both level and composite solves, and a final
synchronization at the end.

@@

Dealing with the energy equation is more difficult. The central
challenge is to write down a form of the discretized energy equation
that explicitly conserves total energy when coupled to gravity. When
gravity is included, the total energy is
\begin{equation}
  \rho E_{\text{tot}} = \rho E - \frac{1}{2}\rho\Phi = \frac{1}{2}\rho \mathbf{u}^2 + \rho e - \frac{1}{2}\rho\Phi, \label{eq:total_energy_gravity}
\end{equation}
where $\rho E$ is the total gas energy from the pure hydrodynamics equation,
and the factor of $1/2$ in the gravitational energy term is necessary
for simulations with self-gravity to prevent double-counting of
interactions. The standard approach has been to note that the total energy update
can be written down by taking the dot product of the velocity and the momentum source term,
so that the time-centered energy update is
\begin{equation}
  \left.\Delta (\rho E)\right|_\text{grav} = \frac{\Delta t}{2} \left[ \rho^n \mathbf{u}^n \cdot \mathbf{g}^n + \rho^{n+1} \mathbf{u}^{n+1} \cdot \mathbf{g}^{n+1}\right].
\end{equation}
However, this does not conserve the total energy of Equation \ref{eq:total_energy_gravity}.
In particular, we notice in our simulations a long-term secular change in the gravitational potential energy
that is not accounted for in the gas energy. One solution to this problem has been offered by 
\cite{jiang:2013} and \cite{marcello:2012}. They propose to evolve the total energy instead of the gas energy 
in the hydrodynamics equations. When this is done, one can write down an explicitly conservative 
formulation:
\begin{equation}
  \frac{\partial \rho E_{\text{tot}}}{\partial t} = -\bm{\nabla}\cdot(\rho\mathbf{u}E + p\mathbf{u} + \mathbf{F_g}).
\end{equation}
Here $\mathbf{F_g}$ is a flux term which explicitly accounts for the change in the total energy due to self-gravity. 
By using Equation \ref{eq:cell_center_gravity_source} and the continuity equation, it is straightforward to show that
\begin{equation}
  {\bm\nabla} \cdot \mathbf{F_g} = -\rho \mathbf{u} \cdot \mathbf{g} - \frac{\partial}{\partial t}\left(\frac{1}{2}\rho\Phi\right). \label{eq:gradient_F_g}
\end{equation}
From here one can use the continuity equation to explicitly calculate the flux $\mathbf{F_g}$, as in \cite{jiang:2013}:
\begin{equation}
  \mathbf{F_g} = \frac{1}{8\pi G}\left(\Phi\bm{\nabla}\dot{\Phi} - \dot{\Phi}\bm{\nabla}\Phi\right) + \rho\mathbf{u}\Phi \label{eq:jiang_flux}.
\end{equation}
Here the dots refer to time derivatives. This has some attractive features. 
In particular, for a simulation with AMR, a flux-based approach
can quite naturally be accommodated in a way that conserves energy across levels, when fluxes
at interfaces between coarse and fine grids are matched. We note in passing that although
\citeauthor{jiang:2013} claim that this approach requires a Poisson solve for the time derivative of the potential,
$\dot{\Phi}$, this is not needed as in practice we will already have the potential at time-levels
$n$ and $n+1$, and to second order in time, $\dot{\Phi}^{n+1/2} = (\Phi^{n+1} - \Phi^{n}) / \Delta t.$

@@

We start with the integral formulation of the energy source term (Equation
\ref{eq:cell_center_gravity_source}), but do not discretize it at
the cell center.  Instead, we recognize that $\mathbf{u} = {\bm\nabla}\left[\mathbf{r}\cdot\mathbf{u}\right]$,
where $\mathbf{r}$ is the position vector originating from the zone center.
Then, since in the cell-centered approach we assume that the density, velocity,
and gravitational acceleration are constant across the zone, we can include all terms
inside the gradient operator. This allows us to use the divergence theorem to evaluate
the quantities on the zone faces. To be second-order accurate in time, the fields must 
all be evaluated at time-level $n+1/2$, but for clarity we will neglect the superscript.
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{\Delta t}{\Delta V} \int \rho \mathbf{u}\cdot\mathbf{g}\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int \rho\, {\bm \nabla} \left[\mathbf{r} \cdot \mathbf{u}\right]\cdot\mathbf{g}\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int {\bm \nabla} \cdot \left[ \rho\, (\mathbf{r} \cdot \mathbf{u})\, \mathbf{g}\right]\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int \rho\, (\mathbf{r} \cdot \mathbf{u})\, \mathbf{g} \cdot \mathbf{dS}_i.
\end{align}
Here $\mathbf{dS}$ is a surface element whose vector points normal to
and away from the surface. In one dimension, this expression evaluates to
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{\Delta t}{\Delta V} \rho_{i-1/2}\, \frac{\Delta x}{2} u_{x_{i-1/2}} g_{x_{i-1/2}} \Delta A_{i-1/2} \notag \\
             &+ \frac{\Delta t}{\Delta V} \rho_{i+1/2}\, \frac{\Delta x}{2} u_{x_{i+1/2}} g_{x_{i+1/2}} \Delta A_{i+1/2} \notag \\
             &= \frac{1}{2} \rho_{i-1/2} u_{x_{i-1/2}} g_{x_{i-1/2}} \notag \\
             &+ \frac{1}{2} \rho_{i+1/2} u_{x_{i+1/2}} g_{x_{i+1/2}}.\label{eq:edge_centered_gravity_source}
\end{align}
(We note that this differs slightly from Equation 94 in \cite{arepo} in that
we evaluate the gravitational acceleration at the zone face rather than the zone center.)
In the limit of constant fields, this expression simply averages out to the standard 
source term at the zone center. However, when the fields vary this can give very different results.

As pointed out by \citeauthor{arepo}, $(\rho u)$ evaluated at the faces at time level $n+1/2$ is the mass flux,
so $(\rho u)_{i-1/2}\, \Delta t\, \Delta A = \Delta m_{i-1/2}$ is the mass exchanged between zones $i$ and $i-1$.
Additionally, to second order $g_{x_{i-1/2}} \Delta x = \Phi_{i} - \Phi_{i-1}$. Therefore the energy gained by the zone is
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{1}{2} \Delta m_{i-1/2} \left(\Phi_{i} - \Phi_{i-1}\right) \notag \\
     &+ \frac{1}{2} \Delta m_{i+1/2} \left(\Phi_{i+1} - \Phi_{i}\right)
\end{align}
This expression says that the energy gained by a zone is due to a mass element moving through a potential gradient
between two zones. Now, the key observation is that this does not have the telescoping property of a conservative flux;
instead, the \textit{same} amount of gravitational potential energy is gained by zones $i$ and $i-1$ by the mass flux 
between them. When this source term is summed up over all zones, it will exactly compensate for the change in 
gravitational potential energy, except for any energy lost through the boundaries. The extension of this energy change 
to multiple dimensions simply means that we include the mass fluxes and potential gradients across the $y$ and $z$ faces.
We have found that in practice this method for conserving the total energy is the least susceptible to deleterious effects
from numerical discretization errors. However, it does mean that there can be sharp changes in the total energy of a zone
in regions near a white dwarf's surface, because the mass fluxes into the ambient medium can be quite large compared to
the mass of the zone. This effect diminishes at higher resolutions as the gradients become better resolved.

@@

Analytic solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. However, on a finite computational domain it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, they can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = \sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{direct_sum}
\end{equation}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,m, n)$ 
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. We have implemented this as an option\footnote{It is controlled
  with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in
CASTRO. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions).

@@

The most natural extension of the monopole prescription is to include
higher-order multipole moments. If the entire mass distribution is
enclosed, then the potential can be expanded in a series of spherical
harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the
coordinate system is taken to be the center of the computational
domain, and $r$ is the distance to the origin. The multipole moments
can be calculated by expanding the Green's function for the Poisson
equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation
\ref{spherical_harmonic_expansion} using the addition theorem for
spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{align}
  \Phi(\mathbf{x}) &= \sum_{l=0}^{\infty} \left\{Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} \right. \notag \\
    &+ \left. \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right\}.\label{multipole_potential}
\end{align}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the
outer summation in Equation \ref{multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions will be directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = \frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}

>>>>

We solve the Poisson equation for self-gravity for our problem,
\begin{equation}
  \nabla^2 \Phi(\mathbf{x}) = 4\pi G\, \rho(\mathbf{x}),
\end{equation}
where $\Phi$ is the gravitational potential, $G$ is the gravitational
constant, and $\rho$ is the mass density.\footnote{In the CASTRO code, the 
right-hand side is negated and therefore $\Phi$ is positive. We use the 
sign convention that is typical for astrophysics in this paper. 
When $\Phi$ appears in the code it is negated to compensate for this.} 
The solution of this equation in CASTRO is described in \cite{castro}, and
consists of both level and composite solves, and a final
synchronization at the end.

@@

Dealing with the energy equation is more difficult. The central
challenge is to write down a form of the discretized energy equation
that explicitly conserves total energy when coupled to gravity. When
gravity is included, the total energy is
\begin{equation}
  \rho E_{\text{tot}} = \rho E + \frac{1}{2}\rho\Phi = \frac{1}{2}\rho \mathbf{u}^2 + \rho e + \frac{1}{2}\rho\Phi, \label{eq:total_energy_gravity}
\end{equation}
where $\rho E$ is the total gas energy from the pure hydrodynamics equation,
and the factor of $1/2$ in the gravitational energy term is necessary
for simulations with self-gravity to prevent double-counting of
interactions. The standard approach has been to note that the total energy update
can be written down by taking the dot product of the velocity and the momentum source term,
so that the time-centered energy update is
\begin{equation}
  \left.\Delta (\rho E)\right|_\text{grav} = \frac{\Delta t}{2} \left[ \rho^n \mathbf{u}^n \cdot \mathbf{g}^n + \rho^{n+1} \mathbf{u}^{n+1} \cdot \mathbf{g}^{n+1}\right].
\end{equation}
However, this does not conserve the total energy of Equation \ref{eq:total_energy_gravity}.
In particular, we notice in our simulations a long-term secular change in the gravitational potential energy
that is not accounted for in the gas energy. One solution to this problem has been offered by 
\cite{jiang:2013} and \cite{marcello:2012}. They propose to evolve the total energy instead of the gas energy 
in the hydrodynamics equations. When this is done, one can write down an explicitly conservative 
formulation:
\begin{equation}
  \frac{\partial \rho E_{\text{tot}}}{\partial t} = -\bm{\nabla}\cdot(\rho\mathbf{u}E + p\mathbf{u} + \mathbf{F_g}).
\end{equation}
Here $\mathbf{F_g}$ is a flux term which explicitly accounts for the change in the total energy due to self-gravity. 
By using Equation \ref{eq:cell_center_gravity_source} and the continuity equation, it is straightforward to show that
\begin{equation}
  {\bm\nabla} \cdot \mathbf{F_g} = -\rho \mathbf{u} \cdot \mathbf{g} - \frac{\partial}{\partial t}\left(\frac{1}{2}\rho\Phi\right). \label{eq:gradient_F_g}
\end{equation}
From here one can use the continuity equation to explicitly calculate the flux $\mathbf{F_g}$, as in \cite{jiang:2013}:
\begin{equation}
  \mathbf{F_g} = \frac{1}{8\pi G}\left(\Phi\bm{\nabla}\dot{\Phi} - \dot{\Phi}\bm{\nabla}\Phi\right) + \rho\mathbf{u}\Phi \label{eq:jiang_flux}.
\end{equation}
Here the dots refer to time derivatives. This has some attractive features. 
In particular, for a simulation with AMR, a flux-based approach
can quite naturally be accommodated in a way that conserves energy across levels, when fluxes
at interfaces between coarse and fine grids are matched. We note in passing that although
\citeauthor{jiang:2013} claim that this approach requires a Poisson solve for the time derivative of the potential,
$\dot{\Phi}$, this is not needed as in practice we will already have the potential at time-levels
$n$ and $n+1$, and to second order in time, $\dot{\Phi}^{n+1/2} = (\Phi^{n+1} - \Phi^{n}) / \Delta t.$

@@

We start with the integral formulation of the energy source term (Equation
\ref{eq:cell_center_gravity_source}), but do not discretize it at
the cell center.  Instead, we recognize that in the limit that the velocity is 
constant, $\mathbf{u} = {\bm\nabla}\left[\mathbf{r}\cdot\mathbf{u}\right]$,
where $\mathbf{r}$ is the position vector originating from the zone center.
Then, since in the cell-centered approach we assume that the density, velocity,
and gravitational acceleration are constant across the zone, we can include all terms
inside the gradient operator. This allows us to use the divergence theorem to evaluate
the quantities on the zone faces. To be second-order accurate in time, the fields must 
all be evaluated at time-level $n+1/2$, but for clarity we will neglect the superscript.
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{\Delta t}{\Delta V} \int \rho \mathbf{u}\cdot\mathbf{g}\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int \rho\, {\bm \nabla} \left[\mathbf{r} \cdot \mathbf{u}\right]\cdot\mathbf{g}\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int {\bm \nabla} \cdot \left[ \rho\, (\mathbf{r} \cdot \mathbf{u})\, \mathbf{g}\right]\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int \rho\, (\mathbf{r} \cdot \mathbf{u})\, \mathbf{g} \cdot \mathbf{dS}_i.
\end{align}
Here $\mathbf{dS}$ is a surface element whose vector points normal to
and away from the surface. In one dimension, this expression evaluates to
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{\Delta t}{\Delta V} \rho_{i-1/2}\, \frac{\Delta x}{2} u_{x_{i-1/2}} g_{x_{i-1/2}} \Delta A_{i-1/2} \notag \\
             &+ \frac{\Delta t}{\Delta V} \rho_{i+1/2}\, \frac{\Delta x}{2} u_{x_{i+1/2}} g_{x_{i+1/2}} \Delta A_{i+1/2} \notag \\
             &= \frac{1}{2} \rho_{i-1/2} u_{x_{i-1/2}} g_{x_{i-1/2}} \notag \\
             &+ \frac{1}{2} \rho_{i+1/2} u_{x_{i+1/2}} g_{x_{i+1/2}}.\label{eq:edge_centered_gravity_source}
\end{align}
(We note that this differs slightly from Equation 94 in \cite{arepo} in that
we evaluate the gravitational acceleration at the zone face rather than the zone center.
We are free to do this since we assumed a uniform gravitational field to begin with.)
In the limit of constant fields, this expression simply averages out to the standard 
source term at the zone center. However, when the fields vary this can give very different results.

As pointed out by \citeauthor{arepo}, $(\rho u)$ evaluated at the faces at time level $n+1/2$ is the mass flux
that is obtained from the Riemann solver in the hydrodynamics step,
so $(\rho u)_{i-1/2}\, \Delta t\, \Delta A = \Delta m_{i-1/2}$ is the mass exchanged between zones $i$ and $i-1$.
Additionally, to second order $g_{x_{i-1/2}} \Delta x = \Phi_{i} - \Phi_{i-1}$. Therefore the energy gained by the zone is
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{1}{2} \Delta m_{i-1/2} \left(\Phi_{i} - \Phi_{i-1}\right) \notag \\
     &+ \frac{1}{2} \Delta m_{i+1/2} \left(\Phi_{i+1} - \Phi_{i}\right)
\end{align}
This expression says that the energy gained by a zone is due to a mass element moving through a potential gradient
between two zones. Now, the key observation is that this does not have the telescoping property of a conservative flux;
instead, the \textit{same} amount of gravitational potential energy is gained by zones $i$ and $i-1$ by the mass flux 
between them. When this source term is summed up over all zones, it will exactly compensate for the change in 
gravitational potential energy, except for any energy lost through the boundaries. The extension of this energy change 
to multiple dimensions simply means that we include the mass fluxes and potential gradients across the $y$ and $z$ faces.
We have found that in practice this method for conserving the total energy is the least susceptible to deleterious effects
from numerical discretization errors. However, it does mean that there can be sharp changes in the total energy of a zone
in regions near a white dwarf's surface, because the mass fluxes into the ambient medium can be quite large compared to
the mass of the zone. This effect diminishes at higher resolutions as the gradients become better resolved.

Our strategy for coupling the gravity to the hydrodynamics on a uniform grid can thus be summarized as follows. 
At the beginning of a timestep, we solve the Poisson equation given the density $\rho^n$. We then solve the 
Euler equations and advance the system in time by $\Delta t$ using zone-centered fluxes at time-level $n+1/2$. 
We add $\Delta t \rho^n \mathbf{g}^n$ to the momentum equations. We then solve the Poisson equation again, given
the density $\rho^{n+1}$. We then add $-(\Delta t/2) \rho^{n} \mathbf{g}^{n} + (\Delta t / 2) \rho^{n+1} \mathbf{g}^{n+1}$
to the momentum equations, time-centering the source term. We simultaneously add $\sum_{j} -(1/2) \Delta m_{j} (\Phi - \Phi_j)$
to the zone's energy, where the index $j$ runs over all adjacent zones. This conserves the total energy on the grid
to roundoff error, modulo any losses from an open domain boundary.

@@

Analytic solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. However, on a finite computational domain it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, they can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = -\sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{direct_sum}
\end{equation}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,m, n)$ 
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. We have implemented this as an option\footnote{It is controlled
  with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in
CASTRO. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions).

@@

The most natural extension of the monopole prescription is to include
higher-order multipole moments. If the entire mass distribution is
enclosed, then the potential can be expanded in a series of spherical
harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = -\sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the
coordinate system is taken to be the center of the computational
domain, and $r$ is the distance to the origin. The multipole moments
can be calculated by expanding the Green's function for the Poisson
equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation
\ref{spherical_harmonic_expansion} using the addition theorem for
spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{align}
  \Phi(\mathbf{x}) &= -\sum_{l=0}^{\infty} \left\{Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} \right. \notag \\
    &+ \left. \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right\}.\label{multipole_potential}
\end{align}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the
outer summation in Equation \ref{multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions will be directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = -\frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}
